### 2004-07-15 Documentation/Paper/paper.tex
Meine Version ...

\vspace*{-3.cm}
\begin{flushright}
TUM-HEP-5XX/04\\
MPP-2004-YY\\
%hep-ph/
\end{flushright}

@@

\begin{abstract}
We present the \GLOBES\ (``General Long Baseline Experiment Simulator'')
software package, which allows the simulation of future long-baseline
neutrino oscillation experiments. One part of the software is the
abstract experiment definition language to define experiments with
beam and full detector simulations as accurate as possible. For the
definition of experiments, a new transparent building block concept is introduced. In addition, a binary provides the possibility to test
new experiment definitions quickly.
%
Another
part of \GLOBES\ is the user's interface, which provides probability, rate, and $\Delta \chi^2$ information for a given experiment or any
combination of up to 32 experiments. Especially, the
$\Delta \chi^2$ functions allow a simulation with statistics only,
systematics, correlations, and degeneracies. In particular, \GLOBES\
can handle the full multi-parameter correlation among the oscillation
parameters, external input, and matter density uncertainties. Eventually,
many systematics options are provided, such as normalization and energy
calibration errors, or the choice between spectral or total rate 
information. 
\end{abstract}

@@

Neutrino oscillations are now established as the leading flavor
transition mechanism for neutrinos. By the existing experiments, 
two oscillation
frequencies have been found: The fast ``atmospheric'' and the
slow ``solar'' oscillations, which are driven by the respective
mass squared differences.  In addition, there could be interference
effects between these two oscillations, provided that the coupling given by the small mixing angle $\theta_{13}$ is large enough. Such interference
effects include, for example, leptonic CP violation.  In order to
test the unknown oscillation parameters, \ie, the mixing angle $\theta_{13}$, the leptonic CP phase, and the
neutrino mass hierarchy, future long-baseline experiments are proposed.
These experiments send an artificial neutrino beam to a detector, or
detect the neutrinos produced by a nuclear fission reactor. However,
instrinsic degenerate solutions in the neutrino oscillation
probabilities affect these measurements and require optimization
strategies for the experiments to maximally exploit complementary
information. Therefore, a modern complete experiment simulation
does not only need to have a beam and detector simulations, which are
as accurate as possible, but also powerful tools to analyze correlations
and degeneracies. In this study, we present the \GLOBES\ software package
as such a tool.

\GLOBES\ (``General Long Baseline Experiment Simulator'') is a flexible
software tool to simulate neutrino oscillation long-baseline experiments using a complete three-flavor analysis. On the
one hand, it contains a comprehensive abstract experiment definition
language, which allows to describe most classes of long baseline experiments
at an abstract level. On the other hand, it provides a C-library to 
process the experiment information in order to obtain oscillation
probabilities, rate vectors, and $\Delta \chi^2$-values (\cf, \figu{GLOBES}). In addition, it provides a binary to test experiment
definitions very quickly, before they are used by the application software.
Currently, \GLOBES\ is available for Linux. Since the source code is included, the modifications to other operating systems should be doable.

\GLOBES\ allows to simulate experiments with stationary neutrino point sources, where each experiment is assumed to have only one neutrino source.
Such experiments are neutrino beam experiments and reactor experiments. 
Geometrical effects of a source distribution, such as in the sun or the 
atmosphere, can not be described. In addition, sources with a physically 
significant time dependencies  can not be studied, such as  supernov\ae. It 
is, however, possible to simulate beams with bunch structure, since the 
time dependence of the neutrino source is physically only important to suppress backgrounds.

@@

On the experiment definition side, either built-in neutrino fluxes
(\eg, neutrino factory) or arbitrary fluxes can be used. Similarly,
arbitrary cross sections, energy dependent efficiencies, the
energy resolution function, the considered oscillation channels, 
backgrounds, and many other features can be specified. 
For the systematics, energy
normalization and calibration errors can be simulated. Note that
the energy ranges and windows, as well as the bin widths can be
(almost) arbitrarily chosen, which means that variable bin
widths are allowed. Together with \GLOBES\ comes a number of
pre-defined experiments in order to demonstrate the capabilities
of \GLOBES\ and to provide prototypes for new experiments.
Examples for these prototypes are the \MINOS , \ICARUS , and
\OPERA\ simulations from \Ref~\cite{Huber:2004ug}, the
\JHFSK\ and \NUMI\ superbeam simulations from \Refs~\cite{Huber:2002mx,
Huber:2002rs}, the \JHFHK\ superbeam upgrade simulation from
\Ref~\cite{Huber:2002mx}, the neutrino factory simulations from
\Ref~\cite{Huber:2002mx}, and the reactor experiment simulations from
\Ref~\cite{Huber:2003pm}. Other projects using the \GLOBES\ 
software include \Refs~\cite{Huber:2003ak,Ohlsson:2003ip,Winter:2003ye,Antusch:2004yx}.

With the C-library, one can extract the $\Delta \chi^2$ for all defined 
oscillation channels for an experiment or any combination of experiments.
Of course, also low-level information, such as oscillation
probabilities or event rates, can be obtained. \GLOBES\ includes the
simulation of neutrino oscillations in matter with arbitrary matter 
density profiles, as well as it allows to simulate the matter density
uncertainty. As one of the most
advanced features of \GLOBES , it provides the technology to 
project the $\Delta \chi^2$, which is a function of all oscillation
parameters, onto any subspace of parameters by local minimization. 
This approach allows the inclusion of multi-parameter-correlations,
where external input (\eg, from solar parameters) can be imposed, too.
Applications of the projection mechanism include the projections onto the $\stheta$-axis and the $\stheta$-$\deltacp$-plane. In addition, all oscillation parameters can be kept free to precisely localize 
degenerate solutions.

@@

In this section, we sketch the computation of 
the event rates, which is on of the core parts of the \GLOBES\
software. However, because of the compecity of this issue,
we refer to the \GLOBES\ manual~\cite{Manual} for
details.

@@

Since it is a lot of effort to solve this double integral numerically,
we split up the two integrations. First, we evaluate the integral over
$\hat{E}$, where the only terms containing $\hat{E}$ are
$k_\beta^{\text{IT}}(E-\hat{E})$,  $ T_\beta(\hat{E})$, and 
$ V_\beta(\hat{E}-E')$. We define:
\begin{eqnarray}
\label{equ:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int\limits_0^\infty d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,. 
\end{eqnarray}
Thus, $R_\beta^{\text{IT}}(E,E')$ describes the energy response of 
the detector, \ie , a neutrino with a (true) energy $E$ is reconstructed
with an energy between $E'$ and $E'+dE'$ with a probability
$R_\beta^{\text{IT}}(E,E') dE'$. The function $R(E,E')$ is also often called ``energy resolution function''. Actually, its internal representation
in the software is a smearing matrix. The function $\epsilon_\beta^{\text{IT}}(E')$ will be refered to as ``post-smearing efficiencies'', since it will allow us to define cuts and threshold functions {\em after} the smearing is performed, \ie, as function of $E'$. 
In addition, \GLOBES\ uses ``pre-smearing efficiencies'', which are 
evaluated {\em before} the smearing is performed, \ie, as function of $E$.
Similarly, constant event rates can be added to the event rates before or after the energy smearing, which are refered to as ``pre-smearing backgrounds'' and ``post-smearing beackground''. These types of 
efficiencies an backgrounds allow a very accurate modeling of
many factors, such as geoneutrino backgrounds, energy cuts, or
energy threshold functions.

@@

\section{Definition of experiments with \AEDL\ (Abstract Experiment 
Defintion Language)}

@@

In order to define experiments, \GLOBES\ uses \AEDL\ (``Abstract
Experiment Definition Language''). An experiment normally corresponds
to an \AEDL\ file, which is a text file written in \AEDL\ syntax.

The main parts of any \AEDL\ experiment definition are channels,
rules, and experiments (\cf, \figu{aedl}). A channel corresponds to a neutrino oscillation channel including flux, cross section (for one specific interaction type), energy resolution function, initial and final neutrino flavors, their polarity (neutrinos or antineutrinos), and efficiencies. The channel raw event rates are computated according to \equ{events_bin}. Each channel leads to the raw event rates for a 
specfifc interaction type. In \AEDL , many of the channel components
have to be defined or loaded from files before, such as fluxes, 
cross sections, or the energy resolution function. For the
different options, see the \GLOBES\ manual~\cite{Manual}.

For a ``rule'', the raw event rates of one or more signal channels and
one or more background channels are added. The splitting in signal and
background is arbitrary, but all of the signal and background components
are defined to have a common systematics. The event rates of all 
signal and background components add up to the total event rate of the
rule, which leads to a $(\Delta \chi^2)_r$. The signal or background
within each rule allows the specification of signal and background normalization errors and energy tilt or calibration errors. 
These systematical errors will be evaluated with the ``pull method''
with the help of nuisance parameters REFERENCE. In addition to these
systematics errors, an overall evaluation strategy is assigned 
to each rule, which specifies the type of systematics (tilt or calibration
error), and the use of spectral information or total event rates.

Finally, one ore more rules add up to an experiment, where the
total $\Delta \chi^2$ is obtained as the sum of the $\Delta \chi^2$
of all rules. This approach allows the definition of 
appearance and disappearance channels, neutrino and antineutrino running, or interaction types with different systematics (spectral information
versus counting rate) within one experiment. The \GLOBES\ user interface
allows the simulation of one or more experiment simultaneously, which
means that one could also use different experiments for different
oscillation channels. However, there is still one component missing
in the experiment definition, which is the matter density profile.
For an experiment, an arbitrary matter density profiles can be specified, which is evaluated with the evolution operator method REFERENCES.
In addition, the matter density profile is multiplied by the matter
density scaling factor, which is treated as an independent parameter
with a relative (matter density) uncertainty. With this
approach, one can take into account that the matter density profile along a specific baseline is only known to about $3\%-5\%$. Therefore, it makes
sense to combine the rules of one baseline in one experiment,
since for all of these rules the matter density profile is equal.

\section{Simulation of experiments: The C user's interface}

@@

In order to use \GLOBES , a C~library is linked with the application
software. This library provides the user's interface functions.
It allows to load \AEDL\ files, initialize the experiments, and
have access to various $\Delta \chi^2$-functions including any combination
of systematics and correlations. In addition, it provides low-level
access to oscillation probabilities and event rates, and allows
the readout and modification of many experiment parameters at running time.

The most sophisticated feature of the user's interface is the
possibility to include the full multi-parameter correlation among
all oscillation parameters. Together with the matter densities
of the $N$ loaded experiments ($N \ge 1$), the neutrino oscillation parameter space in \GLOBES\ has $6+N$ dimensions. In the $\chi^2$-approach,
the allowed region of $k$ ($1\le k<6+N$) fit parameters (at the chosen confidence level) is obtained as the projection of the $6+N$-dimensional fit manifold onto the $k$-dimensional hyperplane. For example,
for the precision of the parameter $\stheta$ only, the fit manifold
needs to be projected onto the $\stheta$-axis. One can easily imagine
that the computation of the full multi-parameter correlation is
very expensive with a grid-based method. However, the topology
of the neutrino oscillation parameter space is well investigated and
rather fortunate. Therefore, we use local minimization for each
solution in the parameter space at the chosen confidence level,
where tools for the location of (disconnected) degenerate solutions 
are provided. For this process, the $6+N$-dimensional local minimizer
is started at the ``guessed'' solution (such as from analytical 
knowledge) and runs into the local minimum. In \figu{projex},
this minimization is illustrated at the simple example of a 
two-parameter correlation at a neutrino factory.

As we have discussed in the last section, the matter density of
each experiment is treated as an addition parameter, which is known
from external measurement with a certain precision. In principle, 
the \GLOBES\ user's interface allows such externally imposed
precisions for all oscillation parameters. This feature turns out
to be especially useful for input from the solar neutrino experiments,
since they provide better precisions for the solar neutrino
osillation parameters $\sdm$ and $\theta_{13}$ than the long-baseline
 experiments themselves.

@@

The running time for the projected $\Delta \chi^2$ is, for a grid based method,
\begin{equation}
 \mathrm{t_{proj}^{Grid}} = \mathcal{O}\left( n^{6+N_{\mathrm{exp}}}\right) \, \times \, \mathrm{t_{sys}},
\end{equation}
where $n$ is the number of steps to be evaluated for each oscillation
parameter, and $N_{\mathrm{exp}}$ is the number of experiments
(since the matter densities of all experiments appear as parameters). 
Though it is, in principle, possible to use such a method with \GLOBES ,
it is impractiable in most cases, since it would involve at least 
billions of
steps for the complete multi-parameter correlation. The local $6+N_{\mathrm{exp}}$-dimensional minimization reduces this effort
to 
\begin{equation}
\mathrm{t_{proj}^{Minimizer}} = \mathcal{O}(1000) \, \times \,
N_{\mathrm{Deg}} \,  \times \, \mathrm{t_{sys}}
\end{equation}
steps, where $N_{\mathrm{Deg}}$ is the number of (disconnected) 
local minima at the chosen confidence level, \ie, the number of
degeneracies.

@@

The \GLOBES\ package~\cite{Globes} is a tar-Ball which consists of the
C-library for the user's interface, a binary to test
\AEDL\ files, example files, 
experiment prototypes with supporting flux and cross 
section files, and other 
supporting files for installation and compilation.
Currently, the C-library is only provided for Linux.
In addition, the user's interface and \AEDL\ manual~\cite{Manual} 
can be obtained separately. Upon extracting the tar-ball,
the \GLOBES\ user's interface library needs to be configured, 
compiled, and installed as 
described in the {\tt INSTALL} file. In principle, it can be
compiled as a static library or a dynamically-linked shared library
(which can be installed with superuser rights or only used in one's
home directory). Then it can be linked against any C program
containing the application software, such as the examples
from the manual. \AEDL\ files, such as the experiment protoypes,
can be edited with any text editor, in order to be loaded by the
users's interface later. In addition, a binary is provided to
test and create experiment descriptions. For example, event rate
information can be quickly provided to adjust the neutrino flux
normalization. For further information, we refer
at this place to the \GLOBES\ manual~\cite{Manual}.

@@

In summary, the \GLOBES\ software package provides powerful tools
to do a full three-flavor analysis of future long-baseline experiments
including systematics, correlations, and degeneracies. In addition,
the abstract experiment definition language allows to define
experiments at an abstract level as accurate as possible. 
Some of the major strengths of \GLOBES\ are the ability to quickly
define new experiments, the potential to take into account the
full multi-parameter correlation, and the possibilities to include
external input and matter density uncertainties.

@@

@@

We would like to thank Martin Freund and Andrea Romanino ???
for initiating this project many years ago with one of the first
neutrino factory simulations. In addition, we would like to thank
Marc Rolinec for the help to translate the experiment descriptions
into \AEDL , and for creating the illustrations in this manuscript.
Finally, thanks to all the people who have been pushing this project
for many years, to the ones who have been continuing asking for the publication of the software, and the referees of several of our
papers for suggestions which lead to improvements in the software.

>>>>

\vspace*{-3.cm}
\begin{flushright}
TUM-HEP-5XX/04\\
%hep-ph/
\end{flushright}

@@

\begin{abstract}
We present the \GLOBES\ (``General Long Baseline Experiment Simulator'')
software package, which allows the simulation of long-baseline and reactor
neutrino oscillation experiments. One part of the software is the
abstract experiment definition language to define experiments with
beam and full detector descriptions as accurate as possible. For the
definition of experiments, a new transparent building block concept is 
introduced. In addition, an additional program provides the possibility 
to develop and test new experiment definitions quickly.
%
Another
part of \GLOBES\ is the user's interface, which provides probability, rate, 
and $\Delta \chi^2$ information for a given experiment or any
combination of up to 32 experiments in C. Especially, the
$\Delta \chi^2$ functions allow a simulation with statistics only,
systematics, correlations, and degeneracies. In particular, \GLOBES\
can handle the full multi-parameter correlation among the oscillation
parameters, external input, and matter density uncertainties. Eventually,
many systematics options are provided, such as normalization and energy
calibration errors, or the choice between spectral or total rate 
information. 
\end{abstract}

@@

Neutrino oscillations are now established as the leading flavor
transition mechanism for neutrinos. In a long history of many experiments, see
\eg~\cite{Barger:2003qi}, 
two oscillation
frequencies have been identified: The fast ``atmospheric'' and the
slow ``solar'' oscillations, which are driven by the respective
mass squared differences.  In addition, there could be interference
effects between these two oscillations, provided that the coupling given by 
the small mixing angle $\theta_{13}$ is large enough. Such interference
effects include, for example, leptonic CP violation.  In order to
test the unknown oscillation parameters, \ie, the mixing angle $\theta_{13}$,
 the leptonic CP phase, and the
neutrino mass hierarchy, new long-baseline and reactor experiments 
are proposed.
These experiments send an artificial neutrino beam to a detector, or
detect the neutrinos produced by a nuclear fission reactor. However,
the presence of multiple solutions which are intrinsic to
 neutrino oscillation probabilities~\cite{Fogli:1996pv,
Burguet-Castell:2001ez,Minakata:2001qm,Barger:2001yr} 
affect these measurements.
Thus optimization strategies are required which maximally exploit 
complementarity between experiments. Therefore, a modern complete experiment 
simulation  and analysis tool does not only need to have 
a highly accurate beam and detector simulation but also powerful 
means to analyze correlations and degeneracies, especially for the combination
of several experiments.
In this paper, we present the \GLOBES\ software package
as such a tool.

\GLOBES\ (``General Long Baseline Experiment Simulator'') is a flexible
software tool to simulate and analyze neutrino oscillation 
long-baseline and reactor experiments using a 
complete three-flavor description. On the
one hand, it contains a comprehensive abstract experiment definition
language (AEDL), which allows to describe 
most classes of long baseline and reactor experiments
at an abstract level. On the other hand, it provides a C-library to 
process the experiment information in order to obtain oscillation
probabilities, rate vectors, and $\Delta \chi^2$-values (\cf, \figu{GLOBES}). 
In addition, it provides a binary program to test experiment
definitions very quickly, before they are used by the application software.
Currently, \GLOBES\ is available for GNU/Linux. Since the source code is 
included, the modifications to other operating systems should be doable.

\GLOBES\ allows to simulate experiments with stationary neutrino point 
sources, where each experiment is assumed to have only one neutrino source.
Such experiments are neutrino beam experiments and reactor experiments. 
Geometrical effects of a source distribution, such as in the sun or the 
atmosphere, can not be described. In addition, sources with a physically 
significant time dependencies  can not be studied, such as  supernov\ae.

@@

On the experiment definition side, either built-in neutrino fluxes
(\eg, neutrino factory) or arbitrary, user defined fluxes can be used. 
Similarly,
arbitrary cross sections, energy dependent efficiencies,
energy resolution functions as well as the considered oscillation channels 
backgrounds, and many other properties can be specified. 
For the systematics, energy
normalization and calibration errors can be simulated. Note that
energy ranges and windows and bin widths can be
(almost) arbitrarily chosen, including bins of different
widths. Together with \GLOBES\ comes a number of
pre-defined experiments in order to demonstrate the capabilities
of \GLOBES\ and to provide prototypes for new experiments.
Examples for these prototypes are the \MINOS , \ICARUS , and
\OPERA\ simulations from \Ref~\cite{Huber:2004ug}, the
\JHFSK\ and \NUMI\ superbeam simulations from \Refs~\cite{Huber:2002mx,
Huber:2002rs}, the \JHFHK\ superbeam upgrade simulation from
\Ref~\cite{Huber:2002mx}, the neutrino factory simulations from
\Ref~\cite{Huber:2002mx}, and the reactor experiment simulations from
\Ref~\cite{Huber:2003pm}. Other projects using the \GLOBES\ 
software include \Refs~\cite{Huber:2003ak,Ohlsson:2003ip,Winter:2003ye,Antusch:2004yx}.

With the C-library, one can extract the $\Delta \chi^2$ for all defined 
oscillation channels for an experiment or any combination of experiments.
Of course, also low-level information, such as oscillation
probabilities or event rates, can be obtained. \GLOBES\ includes the
simulation of neutrino oscillations in matter with arbitrary matter 
density profiles, as well as it allows to simulate the matter density
uncertainty. As one of the most
advanced features of \GLOBES , it provides the technology to 
project the $\Delta \chi^2$, which is a function of all oscillation
parameters, onto any subspace of parameters by local minimization. 
This approach allows the inclusion of multi-parameter-correlations,
where external constraints (\eg, on the solar parameters) can be imposed, too.
Applications of the projection mechanism include the projections onto 
the $\stheta$-axis and the $\stheta$-$\deltacp$-plane. In addition, 
all oscillation parameters can be kept free to precisely localize 
degenerate solutions.

@@

In this section, we sketch the computation of 
the event rates, which is on of the core parts of the \GLOBES\
software. However, because of the complexity of this issue,
we refer to the \GLOBES\ manual~\cite{Manual} for
details.

@@

Since it is a lot of effort to solve this double integral numerically,
we split up the two integrations. First, we evaluate the integral over
$\hat{E}$, where the only terms containing $\hat{E}$ are
$k_\beta^{\text{IT}}(E-\hat{E})$,  $ T_\beta(\hat{E})$, and 
$ V_\beta(\hat{E}-E')$. We define:
\begin{eqnarray}
\label{equ:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int\limits_0^\infty d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,. 
\end{eqnarray}
Thus, $R_\beta^{\text{IT}}(E,E')$ describes the energy response of 
the detector, \ie , a neutrino with a (true) energy $E$ is reconstructed
with an energy between $E'$ and $E'+dE'$ with a probability
$R_\beta^{\text{IT}}(E,E') dE'$. The function $R(E,E')$ is also often called ``energy resolution function''. Actually, its internal representation
in the software is a smearing matrix. The function $\epsilon_\beta^{\text{IT}}(E')$ will be refered to as ``post-smearing efficiencies'', since it will allow us to define cuts and threshold functions {\em after} the smearing is performed, \ie, as function of $E'$. 
In addition, \GLOBES\ uses ``pre-smearing efficiencies'', which are 
evaluated {\em before} the smearing is performed, \ie, as function of $E$.
Similarly, constant\footnote{with respect to the oscillation parameters, 
not the energy} event rates can be added to the event rates before or 
after the energy smearing, which are refered to as 
``pre-smearing backgrounds'' and ``post-smearing background''. These types of 
efficiencies and backgrounds allow a very accurate modeling of
many factors, such as atmospheric or geoneutrino backgrounds, energy cuts, or
energy threshold functions.

@@

\section{Definition of experiments with \AEDL\\ (Abstract Experiment 
Definition Language)}

@@

In order to define experiments, \GLOBES\ uses \AEDL\ (``Abstract
Experiment Definition Language''). An experiment normally corresponds
to an \AEDL\ file, which is a human readable text file written 
in \AEDL\ syntax.

The key concept of \AEDL\ is to regard a detector as an abstract system
which maps the true properties of a neutrino into the reconstructed properties
of the neutrino which subsequently are used in the fit of the oscillation
parameters. Within \GLOBES\ only the energy and flavour are used as 
observables, which is
sufficient for long baseline and reactor experiments. In  other
experiments, like \eg\ atmospheric experiments, more observables like
direction may have to be considered.

The main parts of any \AEDL\ experiment definition are channels,
rules, and experiments (\cf, \figu{aedl}). A channel corresponds to a neutrino oscillation channel including flux, cross section (for one specific interaction type), energy resolution function, initial and final neutrino flavors, their polarity (neutrinos or antineutrinos), and efficiencies. The channel raw event rates are computed according to \equ{events_bin}. Each channel leads to the raw event rates for a 
specific interaction type. In \AEDL , many of the channel components
have to be defined or loaded from files before, such as fluxes, 
cross sections, or the energy resolution function. For the
different options, see the \GLOBES\ manual~\cite{Manual}.

For a ``rule'', the raw event rates of one or more signal channels and
one or more background channels are added. The splitting in signal and
background is arbitrary, but all of the signal and background components
are defined to have a common systematics. The event rates of all 
signal and background components add up to the total event rate of the
rule, which leads to a $(\Delta \chi^2)_r$. The signal or background
within each rule allows the specification of signal and background 
normalization errors and energy tilt or calibration errors. 
These systematical errors are evaluated with the ``pull method'' like 
in~\cite{Huber:2002mx,Fogli:2002pt}.  
In addition to these
systematics errors, an overall evaluation strategy is assigned 
to each rule, which specifies the type of systematics (tilt or calibration
error), and the use of spectral information or total event rates.

Finally, one ore more rules add up to an experiment, where the
total $\Delta \chi^2$ is obtained as the sum of the $\Delta \chi^2$
of all rules. This approach allows the definition of 
appearance and disappearance channels, neutrino and antineutrino running, 
or interaction types with different systematics (spectral information
versus counting rate) within one experiment. The \GLOBES\ user interface
allows the simulation of one or more experiment simultaneously, which
means that one could also use different experiments for different
oscillation channels. However, there is still one component missing
in the experiment definition, which is the matter density profile.
For an experiment, an arbitrary matter density profiles can be specified, 
which is evaluated with the evolution operator method REFERENCES.
In addition, the matter density profile is multiplied by a scaling factor, 
which is treated as an independent parameter
with a relative (matter density) uncertainty. With this
approach, one can take into account that the matter density profile along a 
specific baseline is only known to about $3\%-5\%$.

\section{Analysis of experiments: The C user's interface}

@@

In order to use \GLOBES , a C-library is linked with the application
software. This library provides the user's interface functions.
It allows to load \AEDL\ files, initialize the experiments, and
have access to various $\Delta \chi^2$-functions including any combination
of systematics and correlations. In addition, it provides low-level
access to oscillation probabilities and event rates, and allows
the readout and modification of many experiment parameters at running time.

The most sophisticated feature of the user's interface is the
possibility to include the full multi-parameter correlation among
all oscillation parameters. Together with the matter densities
of the $N$ loaded experiments ($N \ge 1$), the neutrino oscillation parameter space in \GLOBES\ has $6+N$ dimensions. In the $\chi^2$-approach,
the allowed region of $k$ ($1\le k<6+N$) fit parameters (at the chosen confidence level) is obtained as the projection of the $6+N$-dimensional fit manifold onto the $k$-dimensional hyperplane. For example,
for the precision of the parameter $\stheta$ only, the fit manifold
needs to be projected onto the $\stheta$-axis. One can easily imagine
that the computation of the full multi-parameter correlation is
very expensive with a grid-based method. However, the topology
of the neutrino oscillation parameter space is well investigated and
rather smooth. Therefore, it is feasible to use local minimization for each
solution in the parameter space.  For this process, 
a $6+N$-dimensional local minimizer
is started at the ``guessed'' position for the solution
(such as from analytical knowledge) and then runs into a
local minimum. This exercise has to be repeated till all local 
minima, \ie\ degenerate solutions, are found. In \figu{projex},
this minimization is illustrated at the simple example of a 
two-parameter correlation at a neutrino factory.

As we have discussed in the last section, the matter density of
each experiment is treated as an additional parameter, which is known
from external measurement with a certain precision. In principle, 
the \GLOBES\ user's interface allows such externally imposed
precisions for all oscillation parameters. This feature turns out
to be especially useful for input from the solar neutrino experiments,
since they provide better precisions for the solar neutrino
oscillation parameters $\sdm$ and $\theta_{12}$ than long-baseline
 experiments themselves.

@@

The running time for the projected $\Delta \chi^2$ is, for a grid based method,
\begin{equation}
 \mathrm{t_{proj}^{Grid}} = \mathcal{O}\left( n^{6+N_{\mathrm{exp}}}\right) \, \times \, \mathrm{t_{sys}},
\end{equation}
where $n$ is the number of steps to be evaluated for each oscillation
parameter, and $N_{\mathrm{exp}}$ is the number of experiments
(since the matter densities of all experiments appear as parameters). 
Though it is, in principle, possible to use such a method with \GLOBES ,
it is impracticable in most cases, since it would involve at least 
billions of
steps for the complete multi-parameter correlation. The local $6+N_{\mathrm{exp}}$-dimensional minimization reduces this effort
to 
\begin{equation}
\mathrm{t_{proj}^{Minimizer}} = \mathcal{O}(1000) \, \times \,
N_{\mathrm{Deg}} \,  \times \, \mathrm{t_{sys}}
\end{equation}
steps, where $N_{\mathrm{Deg}}$ is the number of (disconnected) 
local minima, \ie, the number of
degeneracies.

@@

The \GLOBES\ package~\cite{Globes} is a tar-ball which consists of the
source code for the C-library providing the user's interface, the source code
 for a program to test
\AEDL\ files, example C-files illustrating the use of the library, 
experiment prototypes in \AEDL\ with supporting flux and cross 
section files, and the usual 
supporting files for installation and compilation.
Currently, \GLOBES\ is only provided for GNU/Linux.
In addition, an extensive manual covering the user's interface and 
\AEDL~\cite{Manual} 
can be obtained from the web-site. \GLOBES\ is free software and as such 
licensed under the GNU Public License.

The installation of \GLOBES\ is highly automated by the use of the
Autotools family and follows the usual procedures. By default
a shared library {\tt libglobes.so} is installed but 
a static version is available
also. Additionally an executable with name {\tt globes} is installed. 
Provisions have been made to install \GLOBES\ without
root privilege into a user's home directory. During the automatic
configuration process also an example Makefile is produced which can
serve as a template for compiling and linking own applications against
{\tt libglobes.so}. This Makefile can be used to directly compile and
link the examples
from the manual.

\AEDL\ files, such as the experiment prototypes,
can be edited with any text editor, in order to be loaded by the
users's interface later. In addition, the {\tt globes} binary allows to
develop and test experiment descriptions. For example, event rate
information can be quickly provided to adjust the neutrino flux
normalization. For further information, we refer
at this place to the \GLOBES\ manual~\cite{Manual}.

@@

In summary, the \GLOBES\ software package provides powerful tools
to do a full three-flavor analysis of future long-baseline and reactor
 experiments
including systematics, correlations, and degeneracies. In addition,
the abstract experiment definition language allows to define
experiments at an abstract level in a highly accurate fashion. 
Some of the major strengths of \GLOBES\ are the ability to quickly
define new experiments, the potential to take into account the
full multi-parameter correlation, the possibilities to include
external input and matter density uncertainties and last but not least
the ability to jointly analyze several experiments.

@@

For the future it is planed to further extend the flexibility of
\GLOBES\ by providing support for dynamically loadable modules, to
allow completely user defined systematics or non-standard oscillation
probabilities.

@@

We would like to thank Martin Freund, who wrote the very first
version of a three-flavour matter profile treatment many years ago.
PH is especially thankful for the invaluable advice of Thomas Fischbacher on
many design issues. 
In addition, we would like to thank
Mark Rolinec for his help to translate the experiment descriptions
into \AEDL , and for creating the illustrations in this manuscript.
Finally, thanks to all the people who have been pushing this project
for many years, to the ones who have been continuing asking for the 
publication of the software, and the referees of several of our
papers for suggestions which lead to improvements in the software.
Parts of this works have been supported by the ``Sonderforschungsbereich
375 f\"ur Astro-Teilchenphysik der Deutschen Forschungsgemeinschaft''.
